---
title: "A Review of Time series classification algorithms"
output: html_notebook
---

This R markdown presents a brief time series classification algorithms review. 
To explore the algorithms it will be using Earth observation time series data set.
So, all document examples focus on Earth observation data. At first in introduction 
we present general concepts about time series before explore them in Earth Observation context. 


The document examples focus on Earth observation data. Therefore the time series are generated through satellite images over a region of interest. The time series samples used in this review are a set of time series divided into two classes: forest and deforestation. The main idea is to present a list of algorithms and how they can be used in this context.

## Introduction

Time series is a sequence values or observations time-ordered. This data type allow explore the the temporal relationship between information observed. In the code above is presented a single time series to example extracted.

```{r}
samples <- readRDS("../../data/rds/samples_def_filtered.rds")
head(samples)
```




-------

## Dynamic Time Warping 

DTW is set of algorithms use to measure the similarity or dissimilarity between the values of two time series to align them. DTW algorithms allow find the optimal non-linear alignment between reference and query. Reference is the target time series pattern, and the query is the time series pattern that we want compare and align. 


```{r}
samples <- readRDS("../../data/rds/samples_def_filtered.rds")
head(samples)
```
